---
title: "PA1_template"
author: "Serafín Moral García"
date: "16 de marzo de 2017"
output: html_document
---

## Loading and preprocessing the data

In first place, we need to read the data and save it in a data frame

```{r}

data <- read.csv("activity.csv")

```

In this case we are not going to do any preproccesign of the data.

## What is mean total number of steps taken per day?

The next step that we have to do is to calculate, for each day, the total number of steps

```{r}

steps_day <- aggregate(steps ~ date, data, sum)
steps <- steps_day$steps

```

Now we make a histogram of the total number of steps taken each day

```{r}
hist(steps, main = "Total number of steps by day", col="yellow", xlab="Number of Steps")

```

We can now calculate the median and the mean of steps per day easily

```{r}
mean_steps <- mean(steps)
median_steps <- median(steps)

```

The mean of total number of steps taken each day is `r mean_steps`.
The median of total number of steps taken each day is `r median_steps`.

## What is the average daily activity pattern?

The first thing that we have to do in order to answer this question is to calculate the mean of steps by interval throug all the days.

```{r}
steps_interval <- aggregate(steps ~ interval, data, mean)

```

Now we plot the average number of steps per as a function of the interval.

```{r}
steps3 <- steps_interval$steps
interval <- steps_interval$interval
plot(interval, steps3, type="l", xlab="Interval", ylab="Total of Steps", main="Average of Steps per Day by Interval")

```

Now we calculate the 5-minute interval which contains the maximum number of steps
on average across all the days in the dataset. For this purpose we first extract the index of the maximum of steps vector and after we use that index for extract the correspondent interval of interval vector. 

```{r}
index <- which.max(steps3)
interval_max <- interval[index]
```

The 5-minute interval which contains the maximum number of steps
on average across all the days is `r interval_max`

## Imputing missing values

Firstly, we calculate the number of missing values in our data set, i.e, the number of cases that are not complete observations.

```{r}
number_missing_values <- sum(!complete.cases(data))
```

The total number of missing values in our data set is `r number_missing_values`

We can observe for which of the columns in our data frame there are missing values

The total number of missing steps is `r sum(is.na(data$steps))`.
The total number of missing dates is `r sum(is.na(data$date))`.
The total number of missing intervals is `r sum(is.na(data$interval))`.

Therefore, we have to input the missing values for steps. We will do it simply assigning the mean of steps.

```{r}
data_new <- data
data_new$steps[is.na(data_new$steps)] <- mean(data$steps)

```

Now we calculate the total number of steps per day in the same way that we have done in the first question.

```{r}

steps_day2 <- aggregate(steps ~ date, data_new, sum)
steps2 <- steps_day2$steps
```

Now we make the new histogram and plot it beside to the histogram plotted in the first question

```{r}
par(mfrow = c(1,2))

hist(steps, main = "Total number of steps by day", col="yellow", xlab="Number of Steps")

hist(steps2, main = "Steps imputing missing values", col="yellow", xlab="Number of Steps")

```

Now we recalculate the values of mean and median


```{r}
mean_steps2 <- mean(steps2)
median_steps2 <- median(steps2)
difference_means <- mean_steps2 - mean_steps 
difference_medians <- median_steps2 - median_steps 

```

The mean of total number of steps taken each day is `r mean_steps2`.
The median of total number of steps taken each day is `r median_steps2`.

The difference between this mean and the mean calculated in the first question is `r difference_means`.

The difference between this median and the median calculated in the first question is `r difference_medians`.


## Are there differences in activity patterns between weekdays and weekends?

We are going to create a new factor variable that indicates if the date corresponds to a weekday or to a weekend day.

```{r}
weekdays <- c("lunes", "martes", "miércoles", "jueves", "viernes")
new_date <- as.Date(data_new$date)
type_day<- as.factor(ifelse(is.element(weekdays(new_date),weekdays), "Weekday", "Weekend"))
data_new$type_day <- type_day
```

Note: We have written the weekdays in Spanish. (lunes = Monday, martes = Thursday, miercoles = Wednesday, jueves = Thuesday and viernes = Friday)

Now we separate the data into weekdays and weekend days.

```{r}
data_weekdays <- data_new[data_new$type_day == "Weekday",]
data_weekends <- data_new[data_new$type_day == "Weekend",]

```
Now we calculate the average number of steps per interval across all days in the same way that we have done in previous questions, for weekdays as well as weekend days. 

```{r}

steps_interval_weekdays <- aggregate(steps ~ interval,data_weekdays,mean)
steps_interval_weekends <- aggregate(steps ~ interval,data_weekends,mean)

```

Finally, we make two plots showing the average number of steps per interval, for weekdays and for weekend days.

```{r}
interval_weekdays <- steps_interval_weekdays$interval
steps_weekdays <- steps_interval_weekdays$steps
interval_weekends <- steps_interval_weekends$interval
steps_weekends <- steps_interval_weekends$steps

par(mfrow = c(2,1))

plot(x = interval_weekdays,xlab = "Interval", y = steps_weekdays, ylab = "average number of steps" ,type ="l",main = "Number of steps by interval at weekdays")

plot(x = interval_weekends, xlab = "Interval", y = steps_weekends, ylab = "average number of steps" ,type ="l",main = "Number of steps by interval at weekends")

```

